---
output:
  bookdown::pdf_document2:
    includes:
      in_header: "preamble.tex"
    toc: false
    number_sections: false
  documentclass: report
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\maketitle

```{r import_data, cache=TRUE}
source("../data-raw/fetch-data.R")
source("../src/faceoffs/data-processing-scripts/create-features.R")
```

Over the course of a typical hockey game, one could expect that around 50 to 75 faceoffs will be contested.
The outcome is often meaningful in determining which team will first dictate the play in the upcoming sequence.
This obvious statement is probably enough to suggest that—as the old saying goes—each and every faceoff is important in a hockey game.
The main objective of this study is to quantify how much. In other words, in leveraging the data at our disposal, we try to answer the following question:

> **What is at stake when the linesman drops the puck?**

It is reasonable to assume that the answer may vary significantly depending on the context in which a particular faceoff occurs.
A non-exhaustive list of important factors may include the current score, the time remaining on the clock, the zone in which the faceoff is taken, and the strength of play (e.g. power play).
Many will also argue that—even though there is not a single coach who overlooks the importance of winning faceoffs—some game plans are better to make the most of an offensive zone faceoff win, while others are more effective at salvaging a faceoff loss in the defensive zone.

The first step in answering our question objectively is to define a reliable metric to help us guide our reasoning, and eventually come up with conclusive statements.
For the purposes of this study, we will evaluate how the outcome of a faceoff affects the likelihood of scoring or allowing a goal on the ensuing sequence.

#### Outline of the study.
The report is divided in three main parts.
In the first part, we analyze the effect of winning faceoffs on a sequence basis, where a sequence is defined as the time between two successive faceoffs.
In the second part, we raise the bar higher by performing a time series analysis in which we evaluate how the probability of scoring a goal evolves over time after the faceoff.
Finally, in the last section, we perform the two aforementioned analyses on a much larger dataset so as to corroborate our initial findings.

## The overall impact of winning faceoffs {#overall}

The first thing to validate is that winning a faceoff does provide an advantage.
A simple way of doing this is to treat each sequence as an observation and fit a logistic regression with a goal indicator as the response variable.
By including the faceoff outcome as a covariate, we hope to validate that there is statistical evidence that winning a faceoff has a positive effect.
Although reducing the probability of allowing a goal is as important as increasing the probability of scoring a goal, we focus on the latter in this section.

```{r features1}
library(data.table)
library(dplyr)
library(kableExtra)

dt_table <- sequences[, .(
  min_date = min(game_date),
  max_date = max(game_date),
  nb_games = length(unique(game_date)),
  nb_fo = .N,
  nb_fo_oz = sum(faceoff_zone == "offense"),
  nb_goals_for = sum(result_goal_for),
  nb_goals_oz = sum(faceoff_zone == "offense" & result_goal_for)
)]

knitr::kable(
  x = dt_table,
  format = "latex",
  booktabs = T, 
  col.names = c("Min. game date", "Max. game date", "Games", "FO", "OZ FO", "Goals", "OZ Goals"),
  align = "c",
  center = TRUE,
  caption = "High-level features of the scouting dataset (Erie Otters perspective)"
)
```

In order to work with relatively homogeneous data, we made the decision to only use the [scouting dataset](https://raw.githubusercontent.com/bigdatacup/Big-Data-Cup-2021/main/hackathon_scouting.csv).
Indeed, mixing different leagues may introduce noise, as each of them may imply a different style of hockey.
Furthermore, the fact that every games in the data involve the Erie Otters allows us to create models from a single perspective.
It would be somewhat inappropriate to build a model with such asymmetrical data and claim that it represents the reality for the whole league, the OHL in this case.

With that in mind, let us take a look at the data we are working with.
A summary of it is provided in Table \@ref(tab:features1)—OZ stands for Offensive Zone—and an excerpt of the training dataset we use for the logistic regression, including created features, is displayed in Table \@ref(tab:preview).

```{r preview}
sequences[, faceoff_oz := faceoff_zone == "offense"]
columns_show <- c("game_date", "period", "clock_begin", "clock_end", "faceoff_win", "faceoff_zone", "faceoff_oz", "result_goal_for")
knitr::kable(
  x = head(sequences[, .SD, .SDcols = columns_show]),
  booktabs = T,
  col.names = c("game_date", "period", "clock_begin", "clock_end", "FO_win", "FO_zone", "FO_OZ", "GF"),
  align = "c",
  center = TRUE,
  caption = "Excerpt of the sequence-by-sequence data",
  linesep = ""
)
  
```

```{r glm}
mod <- glm(
  formula = result_goal_for / length_seconds ~ faceoff_win,
  family = binomial,
  data = sequences
)

summary <- summary(mod)

mod2 <- glm(
  formula = result_goal_for / length_seconds ~ faceoff_win + faceoff_oz,
  family = binomial,
  data = sequences
)

summary2 <- summary(mod2)

mod3 <- glm(
  formula = result_goal_for / length_seconds ~ faceoff_win * faceoff_oz,
  family = binomial,
  data = sequences
)

summary3 <- summary(mod3)
```

To get a crude idea of the impact on goal scoring of winning or losing faceoffs, we first fitted a logistic regression with only an intercept and `FO_win` as covariate.
We obtained $\beta_{\rm FO\_win} =$ `r round(summary$coefficients[2,][1], 3)`, which suggests that winning offensive faceoffs does increase the chances of scoring a goal during the sequence that ensues.
However, the associated p-value is too high to make any statistically backed conclusion.
We nevertheless designed alternative linear predictors using additional covariates to try to capture a clearer signal.
Adding the binary `FO_OZ` covariate in the model, we obtained two positive coefficients $\beta_{\rm FO\_win} =$ `r round(summary2$coefficients[2,][1], 3)` and $\beta_{\rm FO\_OZ} =$ `r round(summary2$coefficients[3,][1], 3)`.
Since there is probably an important interaction between the faceoff zone and the outcome of the faceoff, we fitted a last logistic model including the `FO_win` * `FO_OZ` interaction.
With this model, the coefficients of interest were $\beta_{\rm FO\_win} =$ `r round(summary3$coefficients[2,][1], 3)`, $\beta_{\rm FO\_OZ} =$ `r round(summary3$coefficients[3,][1], 3)` and $\beta_{\rm FO\_win + FO\_OZ} =$ `r round(summary3$coefficients[4,][1], 3)`, suggesting that winning a faceoff or having an offensive zone faceoff gives great opportunities, but that the real deal is to win those offensive zone faceoffs.
It must be said however that all p-values were well above any reasonable threshold for statistical significance.

We conclude this section with a one-way analysis, shown in Table \@ref(tab:features2), that groups faceoffs by "situation" (penalty kill, even strength, power play).
Although analyzing the impact of winning faceoffs in such specific situations would have been very interesting, Table \@ref(tab:features2) suggests that this would be inconclusive.

```{r features2}
sequences[, goal_strength := "even"]
sequences[team_skaters_nb > opponent_skaters_nb, goal_strength := "pp"]
sequences[team_skaters_nb < opponent_skaters_nb, goal_strength := "pk"]

sequences[faceoff_zone %in% c("blue_offense", "red", "blue_defense"), faceoff_zone := "neutral"]

table_cells <- function(goal, zone, strength) {
  
  goal <- paste0("result_goal_", goal)
  summary <- sequences[goal_strength == strength & faceoff_zone == zone, .N, c("faceoff_win", goal)]
  setorderv(summary, c("faceoff_win", goal), c(-1, 1))
  
  row_1 <- summary[, .(N = sum(N)), .(faceoff_win)][, N]
  row_2 <- summary[get(goal) == TRUE, N]
  row_3 <- row_2 / row_1
  
  row_1 <- as.character(row_1)
  row_2 <- as.character(row_2)
  row_3 <- paste0(as.character(round(row_3 * 100)), "%")
  
  as.matrix(rbindlist(list(
    list(W = row_1[1], L = row_1[2]),
    list(W = row_2[1], L = row_2[2]),
    list(W = row_3[1], L = row_3[2])
  )))
  
}

contextual_data <- cbind(
  rep(c("PK", "Even", "PP"), each = 3L),
  rep(c("Sequence", "Goal", "Rate"), 3L),
  rbind(
    table_cells("against", "defense", "pk"),
    table_cells("against", "defense", "even"),
    matrix(rep("", 6L), ncol = 2L)
  ),
  rbind(
    matrix(rep("", 6L), ncol = 2L),
    table_cells("against", "neutral", "even"),
    matrix(rep("", 6L), ncol = 2L)
  ),
  rbind(
    matrix(rep("", 6L), ncol = 2L),
    table_cells("for", "neutral", "even"),
    matrix(rep("", 6L), ncol = 2L)
  ),
  rbind(
    matrix(rep("", 6L), ncol = 2L),
    table_cells("for", "offense", "even"),
    table_cells("for", "offense", "pp")
  )
)

kbl(
  x = contextual_data,
  format = "latex",
  position = "H",
  caption = "Contextual data for faceoff situations",
  booktabs = TRUE,
  align = "c",
  col.names = c("", "Faceoff", rep(c("Won", "Lost"), 4L))
) %>%
  column_spec(1L, bold = TRUE) %>%
  collapse_rows(columns = 1L, latex_hline = "major", valign = "middle") %>%
  add_header_above(c(" " = 2, "Defense" = 2, "Neutral" = 2, "Neutral" = 2, "Offense" = 2)) %>%
  add_header_above(c(" " = 2, "Goal Against" = 4, "Goal For" = 4), bold = TRUE) %>%
  column_spec(3:10, width = "1cm")
```


## Breaking down the sequences by seconds
We now restrict our attention to offensive zone faceoffs.
So far, we have gathered *some* evidence that winning these increases the chances of scoring a goal.
We know from experience that winning an offensive zone faceoff often leads to a quick shot from the blue line, or more generally to a scoring chance of some sort.
Therefore, our intuition tells us that the effect should be most prominent in the very first seconds following the faceoff and that it should become less and less relevant as time goes by.
To verify that on the Erie dataset, we look
To back our intuition, let us take a closer look at the times the Erie goals in question were scored; for this analysis, we also look at the goals against Erie following a defensive faceoff.
Although we do have the exact goal times (down to the second), we prefer to bin them over windows of 5 seconds for now, as it cancels out some of the noise in the data.

\begin{figure}
    \centering
  \includegraphics[]{"figures/bar_off.png"}
  \qquad 
  \includegraphics[]{"figures/bar_def.png"}
  \caption{Percentage of times a goal was scored within the time window $[t,t+5)$ following the faceoff, {\bf given that the sequence in question lasted at least $t$ seconds}. The left panel is based on goals for Erie following offensive zone faceoffs, while the right panel is based on goals against Erie following defensive zone faceoffs. The data was collected during the 2019-20 season.}
  \label{fig:barcharts}
\end{figure}

Figure \ref{fig:barcharts} is encouraging: it suggests that winning an offensive faceoff (or losing a defensive faceoff) leads more often to a *quick goal*, where by *quick goal* we mean within the first five-ten seconds of the sequence.
It also suggests that the impact of winning the faceoff in offense (green bars in the left panel, and red bars in the right panel) vanishes after approximately 20 seconds. 
However, note that no bar actually correspond to more than ZZZ goals, and that there are some peculiarities to the data, like the fact that the bin $[45,50)$ of the offensive panel shows no goal while its neighbors are significantly different from zero.
Such anomalies are more likely to happen in later bins, as the number of sequences that reaches 45 seconds, say, is much less that those reaching 10 seconds (\textcolor{red}{XX vs YY} in this particular case).
While interpreting Figure \ref{fig:barcharts}, it's important to note that it provides conditional probabilities, and therefore that the numbers shown are theoretically invariant to the performance on the faceoffs themselves; although losing more faceoffs, for example, necessarily implies more data for estimating the quantities of interest for this particular situation.

The noise in the barcharts of Figure \ref{fig:barcharts} caused by the lack of data points makes it a bit hard to interpret them.
To overcome such difficulty, we now construct a smooth version of these latter using a generalized additive model (*gam*).
We relegate the more formal description of the underlying model to the [Technical details section](#technical-details) at the end of the report.
Basically, we wish to construct a curve that provides the *risk* of a goal being scored at any time $t$, conditional on the outcome of the faceoff.
This *risk* can be interpreted as the probability of a goal being scored within a one-second time frame.
To do so, we first expand the sequence-by-sequence data so that each row of the new data corresponds to a one-second window of time, much like we have done for the barcharts, but with bins of the form $[t,t+1)$ this time.
The resulting curves are shown in Figure \ref{fig:gam}.

\begin{figure}
  \centering
  \includegraphics[]{"figures/curve_off.png"}
  \qquad 
  \includegraphics[]{"figures/curve_def.png"}
  \caption{Model-based Erie's goal rates for time windows of the form $[t,t+1)$, \textbf{given that the sequence in question lasted at least $t$ seconds.} The left panel covers goals for following offensive zone faceoffs, while the right panel covers goals against following defensive zone faceoffs. The data was collected during the 2019-20 season.}
  \label{fig:gam}
\end{figure}

Since Figure \ref{fig:gam} pretty much conveys the same information as Figure \ref{fig:barcharts}, it is not surprising that is shows nice bumps right after the faceoff when this latter is won by the offensive team (Erie in the left panel, and its opponent in the right panel).
Its continuous nature makes it easier to compare the offensive and defensive performances of Erie.
For example, they seem to be able to limit their opponents' ability to score when they lose the faceoff in defensive zone (right panel, red curve), as their opponents' performance in similar situation (left panel, green curve) shows a much bigger bump, which corresponds to goals for Erie.
We again note that when the advantage provided by winning an offensive faceoff seems to vanish around 20 seconds into the sequence.
It also seems that when Erie is in defense, winning the faceoff is almost disadvantageous.
This is most probably due to the lack of data, and the variability that ensues.
In fact, it must be said that the results we have so far are not "significant", in the statistical sense.
The pointwise confidence intervals associated to all four curves, also reported in \ref{fig:gam}, are very large, and so any conclusion drawn from them or from the barcharts in Figure \ref{fig:barcharts} must be taken with a grain of salt.
Furthermore, the width of these confidence intervals strongly suggests that performing a similar analysis with a more granular definition of "situation", e.g. distinguishing even strength, powerplay and penalty kill sequences, is probably not a good idea.


## Going bigger

In view of our analysis of the Erie dataset, it seems more than reasonable to think that winning offensive faceoffs does increase one's chances of scoring a goal.
Looking back at the fact that the various probabilities estimated were no greater than 0.5\% = 0.005, it would have been surprising to obtain statistically significant results with the `r dt_table$nb_fo_oz_dz` offensive faceoffs of the Erie dataset.

In order to convince ourselves that our intuition is nevertheless backed by data, we now perform a similar analysis using the 2019-20 NHL data.\footnote{We fetched the data using \href{https://jplecavalier.github.io/tidynhl/}{\texttt{tidynhl}}, an \texttt{R} package developed on top of the NHL Stats and Records APIs.}
We consider this data to be balanced, as all teams appear in it roughly the same amount of times.
Table \@ref(tab:nhl1) provides a basic summary of this NHL dataset.

```{r nhl1}
seasons_to_keep <- c("2019-20")
data_nhl <- fread("../data/nhl-2.csvy", yaml = T)
dt_model_nhl <- data_nhl[season_years %in% seasons_to_keep]

dt_table_nhl <- dt_model_nhl[, .(
  min_date = as.Date(min(game_datetime)),
  max_date = as.Date(max(game_datetime)),
  nb_games = length(unique(game_datetime)),
  nb_fo = .N,
  nb_fo_oz_dz = sum(faceoff_zone == "offense"),
  nb_goals = sum(result_goal_for),
  nb_goals_oz_dz = sum(faceoff_zone == "offense" & result_goal_for)
)]

knitr::kable(
  x = dt_table_nhl,
  format = "latex",
  booktabs = T, 
  col.names = c("Min. game date", "Max. game date", "Games", "FOs", "OZ/DZ FOs", "Goals", "OZ/DZ Goals"),
  align = "c",
  center = TRUE,
  caption = "High-level features of the NHL dataset"
)
```

```{r nhl_glm}
dt_model_nhl[, faceoff_oz := faceoff_zone == "offense"]

mod_nhl <- glm(
  formula = result_goal_for / length_seconds ~ faceoff_win * faceoff_oz,
  family = binomial,
  data = dt_model_nhl
)

summary_nhl <- summary(mod_nhl)
```

#### The overall impact of winning a faceoff.
It seems worthwhile pointing out that we now work with `r dt_table_nhl$nb_fo_oz_dz` faceoffs that lead to `r dt_table_nhl$nb_goals_oz_dz` goals.
As we did in the first section of the report, we fit a logistic regression on this data.
We directly fitted the more complex model with an intercept, `FO_win` and `FO_OZ` as covariates, an interaction between those two covariates, and the variable `GF` as response.
We obtained $\beta_{FO\_win} =$ `r round(summary_nhl$coefficients[2,][1], 3)`, $\beta_{FO\_OZ} =$ `r round(summary_nhl$coefficients[3,][1], 3)` and $\beta_{FO\_win + FO\_OZ} =$ `r round(summary_nhl$coefficients[4,][1], 3)`. This time, due to the much larger dataset, all the p-values but the one of $\beta_{FO\_win}$ were statistically significative.
The conclusion reached in the first section is still valid, there is nothing better than winning an offensive faceoff to increase the probability of scoring a goal.
Furthermore, we can now say that the edge is better if you can get an offensive faceoff—win it or lose it—than if you win a faceoff anywhere else on the ice.

#### Breaking down the sequences by seconds.
We again begin the more granular analysis by constructing a barchart showing the proportion of goals (for) scored in a given time window, conditional on the outcome of the faceoff.
The huge amount of data now allow us to use smaller bins of two seconds, that is $[t,t+2)$.
The result is displayed on the right panel of Figure \ref{fig:barchart-nhl}.

\begin{figure}
  \centering
  \includegraphics[]{"figures/bar_off_nhl.png"}
    \qquad 
  \includegraphics[]{"figures/curve_off_nhl.png"}
  \caption{Right panel: Percentage of times a goal was scored within the time window $[t,t+2)$ following offensive zone faceoffs. Left panel: Model-based goal for rates for time windows of the form $[t,t+1)$. All quantities are {\bf conditional on the fact that the sequence in question lasted at least $t$ seconds.} The results are averaged over the data from all NHL teams collected during the 2019-20 season. The dotted line on the right panel shows the results from an alternative model discussed in the *Technical details* section.}
  \label{fig:barchart-nhl}
\end{figure}

From the Figure \ref{fig:barchart-nhl}, we can see better the effect of winning an offensive zone faceoff, especially within 20 seconds after a faceoff. Let us construct the smoother version of this bar chart using our gam model (see Figure \ref{fig:barchart-nhl}). With this dataset, it is now more clear compared to what we had in the Figure \ref{fig:gam}.

In the Figure \ref{fig:barchart-nhl}, we can see again the bump when winning the faceoff in the moments that follows. When can also see that the confidence intervals are now much more smaller (compare to Figure \ref{fig:gam}), and do not overlap between each others in the first 20 seconds (approximately). Then, we are able to say that the difference is significant for this time. Beyond that 20 seconds mark, we see that both curves converges and are not significantly different. At some point, we can conclude that winning a faceoff does not provide an advantage that lasts forever.

Now that we understand the effect of winning offensive faceoffs, we can ask ourselves, is this effect the same among all teams? To compare the different teams against each others, we focused on the so called "bump" that we see after winning (or loosing) a faceoff. To take into account both the offense and the defense, we compare the heights of both bumps (offensive and defensive) for all teams. The offensive bump can be interpreted as the ability to score a goal after a won offensive faceoff while the defensive bump can be interpreted as the ability to contain the opponent after a lost faceoff. In the Figure XX, we can see the offense on the x axis, and the defense on the y axis. In the best scenario, a team would prefer to have a high bump on offense and a small bump on defense (which corresponds to the bottom right corner).

FIGURE XXX

ANALYSE DE LA FIGURE


\bigskip
\bigskip

#### Let's wrap this up

It's fair to say that most of the people watched enough hockey games could have guessed that winning an offensive zone faceoff increases the chances to score a goal in the sequence that follows. However, this analysis allowed us to conclude few interesting things:

- with enough data, we can conclude that the increase is statistically significant 
- the effect of winning a faceoff does not lasts forever (about 20 seconds)
- COMPARAISON DES ÉQUIPES (CONCLUSION)

\begin{figure}
  \centering
  \includegraphics[]{"figures/xy_nhl.png"}
  \qquad 
  \includegraphics[]{"figures/curve_BOS_BUF.png"}
  \caption{Model-based goal rates for windows of the form $[t,t+1)$, \textbf{given that the sequence in question lasted at least $t$ seconds.} The left panel concerns goals for Erie following offensive zone faceoffs, while the right panel concerns goals against Erie following defensive zone faceoffs. The data was collected during the XXXX season.}
  \label{fig:gam}
\end{figure}

We were able to breakdown the faceoffs in different situations, but we think it would also be interesting to explore or push further other situations:

- compare the effect of winning an offensive zone faceoff in power play compare to even strength
- AUTRE OUVERTURE


\medskip
\rule{\textwidth}{1pt}

## Technical details {#technical-details}

#### Figure \ref{fig:gam} and \ref{fig:barchart-nhl}
For convenience, we focus on offensive faceoffs here.
Denote $Y_t \in \{0,1\}$ the random variable indicating whether a goal (for) occurred in the interval of time $[t,t+1)$, and let $\mathbb{E}(Y_t|x)$ be the expectation of $Y_t$ (i.e., the probability of a goal occurring in that timespan) given that the most recent faceoff was lost ($x = 0$) or won ($x = 1$).
The curves in Figure \ref{fig:gam} were obtained by fitting the *generalized additive model* (*gam*)
\begin{equation} \label{eq:gam1}
  g\{\mathbb{E}(Y_t|x)\} = \beta + (1 - x) f_{0}(t^*) + x f_{1}(t^*), \qquad g(z) = \ln\{z/(1-z)\},
\end{equation}
where $g$ is the so-called logit function and, for both $k=0,1$, $f_k$ is an unknown (i.e., to be estimated) nonlinear function of $t^* = h(t)$ that approximates $g(Y_t|x = k)$. To fit *gam*'s, we use the function \texttt{gam} of the \texttt{R} package \texttt{mgcv}\footnote{More details. We use the \texttt{R} function \texttt{s} to estimate the functions $f_k$, $k=0,1$; in doing so, we specify the option \texttt{pc=0} to force $f_k(0) = 0$ for both $k=0,1$; see GITHUB LINK for the exact code. The package \texttt{mgcv} implements a Bayesian approach.}.
As explained in the main text, we use $t^* = \log(t+1)$ to generate Figure \ref{fig:gam}, so as to favor less smoothing near $t=0$, where we expect a more rapidly changing curve, and more smoothing for larger values of $t$, for which less data is available.
By design, this helps obtaining a (nearly) null risk that a goal occurs at $t = 0$, which is obviously what one would expect.
The probabilities reported are obtained by solving this \eqref{eq:gam1} for $\mathbb{E}(Y_{t^*}|x)$ at each value of $t \in \{0,1,2,\dots\}$, using the fact that $t^* = h(t)$ for some function $h$.
Two models, one using $t^* = log(t+1)$ and another using $t^* = t$ are shown in Figure \ref{fig:barchart-nhl}.



