---
title: "Data analysis"
output:
  html_document:
    df_print: paged
---

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

ggplot2::theme_set(ggplot2::theme_classic())
ggplot2::theme_update(
  
  panel.grid.major.y = ggplot2::element_blank(),
  legend.position = "bottom",
  legend.title = ggplot2::element_text(size = 8),
  axis.title.y = ggplot2::element_blank()
  
)
```

This is a quite exploratory data analysis on the "scouting" data. That dataset consists of Erie Otters 2019-2020 season data.

## Overall data

Let's first import our dataset and look at a couple of high-level things.

```{r import_data}
library(rmarkdown)
source("data-raw/fetch-data.R")
data <- data[dataset == "scouting",]
paged_table(data)
```

How many games are we looking at? 

```{r}
length(unique(data$game_date))
table(data$period)
```

Let's look at the number of observations and also how many of them implies Erie Otters team versus other teams.

```{r}
dim(data)
dim(data[team == "Erie Otters",])
```

Makes sense, almost half of the occurrences implies the Erie Otters team. Considering we have `r length(unique(data$game_date))`, and that a game is 60 minutes, we then have on average `r round(nrow(data)/length(unique(data$game_date)))` ocurrences per game, which correspond to an entry every `r round((60*60)/(nrow(data)/length(unique(data$game_date))))` seconds. Wow ! Let's see the distribution of the number of events per second:

```{r}
library(ggplot2)
nb_events <- data[, .(nb_events = .N), by = .(game_date, period, clock)]
hist(nb_events$nb_events)
table(nb_events$nb_events)
wow <- nb_events[nb_events == 12,]
data[game_date == wow$game_date & period == wow$period & clock == wow$clock,]
```

Overall, most of the moment contains 1 or 2 events ... By looking at the extermes we noticed it's because of penalties (like major fighting haha).

One of the most important feature of this dataset is the `Event`, that basically says the row correspond to what kind of entry.

```{r}
table(data$event)
round(table(data$event)/nrow(data), 2)
```

So we see a lot of `Play`, which corresponds to Pass attempts that are successful. Fortunately there is more successful pass than unsucessful (`Incomplete Play`) ... 

## Faceoffs

Because we want to put the emphasis on the faceoffs events for now, let's have a closer look at these kind of events.

Let's look at the overall faceoff win % of Erie Otters and the number of faceoffs

```{r}
nrow(data[event == "Faceoff Win",])
nrow(data[event == "Faceoff Win" & team == "Erie Otters",])/nrow(data[event == "Faceoff Win",])
```

Not that good ... let's look at the best players of the Erie Otters.

```{r}
fo_win <- data[event == "Faceoff Win" & team == "Erie Otters", .(nb_fo_won = .N), by = .(player)]
fo_lost <- data[event == "Faceoff Win" & team != "Erie Otters", .(nb_fo_lost = .N), by = .(player = player_2)]
fo_stats <- merge(fo_win, fo_lost, by = "player", all.x = TRUE)
fo_stats[is.na(nb_fo_won), nb_fo_won := 0]
fo_stats[is.na(nb_fo_lost), nb_fo_lost := 0]
fo_stats[, nb_fo_total := nb_fo_won + nb_fo_lost]
fo_stats[, fo_perc := nb_fo_won/nb_fo_total]
fo_stats[order(-fo_perc)]
```

Let's look at the faceoff taken in power play versus even stength,

```{r}
data[away_team_skaters == home_team_skaters, pp_situation := "even"]
data[home_team == "Erie Otters" & home_team_skaters > away_team_skaters, pp_situation := "pp_erie"]
data[home_team != "Erie Otters" & home_team_skaters < away_team_skaters, pp_situation := "pp_erie"]
data[home_team == "Erie Otters" & home_team_skaters < away_team_skaters, pp_situation := "pk_erie"]
data[home_team != "Erie Otters" & home_team_skaters > away_team_skaters, pp_situation := "pk_erie"]
data[event == "Faceoff Win" & team == "Erie Otters", fo_won_erie := TRUE]
data[event == "Faceoff Win" & team != "Erie Otters", fo_won_erie := FALSE]
table(data[event == "Faceoff Win"]$pp_situation)

fo_data <- data[event == "Faceoff Win", .(fo = .N), .(pp_situation, fo_won_erie)]
fo_data_tot <- data[event == "Faceoff Win", .(total_fo = .N), .(pp_situation)]
fo_data <- merge(fo_data, fo_data_tot, by = "pp_situation", all.x = TRUE)
fo_data[, fo_perc := fo/total_fo]

ggplot(fo_data, aes(x = pp_situation, y = fo, fill = fo_won_erie)) +
  geom_col(position = "fill") +
  geom_text(data = fo_data[fo_won_erie == TRUE], aes(x = pp_situation, label = round(fo_perc, 2), y=fo_perc + 0.04))
```

Now, let's look at the position of the faceoff on the ice: 

```{r}
table(data[event == "Faceoff Win",]$x_coordinate)
sum(table(data[event == "Faceoff Win",]$x_coordinate))
ERIE_ZONES <- data.table(
  period = rep(c(1:4), 2),
  x_coordinate = c(169, 31, 169, 169, 31, 169, 31, 31),
  fo_zone = c("off", "def", "off", "def", "off", "def")
)
data_fo_position <- merge(data[event == "Faceoff Win",], ERIE_ZONES, by = c("period", "x_coordinate"), all.x = TRUE)
data_fo_position[event == "Faceoff Win" & is.na(fo_zone), fo_zone := "neutral"]
table(data_fo_position$fo_zone)
```

In power play only:

```{r}
table(data_fo_position[pp_situation == "pp_erie",]$fo_zone)
round(table(data_fo_position[pp_situation == "pp_erie",]$fo_zone)/nrow(data_fo_position[pp_situation == "pp_erie",]), 2)
```

